### 2020-07-10 paper/sex_selection_spacing-ver4.tex
Clarify spell definition; use birth interval rather than spell length

First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell, as is assumed by the proportional hazard model.
If proportionality does not hold, the results are biased.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.

The introduction of prenatal sex determination exacerbates any bias from
the proportionality assumption for two reasons. 
First, sex-selective abortion use varies across groups, which affects birth spacing. 
Second,
a household's use of sex selection may vary within a spell, which means
that the effects of covariates vary within the spell as well.

I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

The model is a discrete-time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from nine months after one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise constant baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2).
\end{equation}
$D_j(t)$ is the piece-wise constant baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, and $Z_2$ captures the area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.%
\footnote{
With sex selection, the composition of prior children is, in principle, endogenous.
It is beyond the scope of this paper to develop a method for dealing with this issue.
}
However, to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.%
\footnote{
I discuss the choice of which variables to use for non-proportionality in more detail
in the ``Explanatory Variables'' section.
}

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is an interval---here equal
to three months---I can estimate the model using a standard multinomial logit model.
%
% \footnote{
% The multinomial model assumes that alternative exit states are stochastically independent
% (Independence of Irrelevant Alternatives or IIA), which rules out any individual-specific 
% unmeasured or unobservable factors that affect both the hazard of having a girl and the 
% hazard of having a boy.
% I, therefore, include a proxy for fecundity discussed in Section \ref{sec:data}.
% Also, the multivariate probit model can be used as an alternative because the IIA is not 
% imposed \citep{han90}.
% The results are substantially identical between these two models and
% available upon request.
% }
In the reorganized data the outcome variable is 0 if the woman does not have a child in a 
given interval (the base outcome), 1 if she gives birth to a son in that interval, and 2 
if she gives birth to a daughter in that interval.

@@

The interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy, $b$, and of having a girl, $g$, in period $t$, conditional on 
a set of explanatory variables and not having had a child before that period.
% , as
% \begin{align}
% P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
% {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})}
% \label{eq:probability_boy} \\
% P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
% {1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}
% \label{eq:probability_girl}
% \end{align}
It is then straightforward to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage of boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent of births that are boys over the entire spell.

@@

The data come from the four rounds of the National Family Health Survey
collected in 1992--1993, 1998--1999, 2005--2006, and 2015--2016.
The surveys are large: 89,777, 90,303, 124,385, and 699,686 women,
respectively. NFHS-1 and NFHS-2 surveyed only ever-married women, while
the later surveys also included never-married.

I exclude visitors, as well as women in any of the following 
categories: never married; no gauna yet; married more than once; divorced; 
not living with husband; inconsistent age at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported giving birth 
before age 12, had a birth before marriage, or an interval between births of less than 
nine months.

Finally, I restrict the sample to Hindus, who constitute about 80 percent of India’s 
population.
If the use of sex selection differs across groups, assuming that the
baseline hazard is the same leads to bias. 
Furthermore, the other groups are each too small to estimate different
baseline hazards for each and so different that combining them into one
group would not make sense.

In addition to a large number of women surveyed and the long period covered, a
benefit of the NFHS is that enumerators pay careful attention to the spacing 
between births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, remains a potential problem.
Recall error is heavily dependent on how long ago a woman was married, and I, therefore, 
drop women married 22 years or more based on the discussion in the Appendix.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

I focus on the second through fourth spells, that is, on the intervals from the first 
birth until at most the fourth birth.
I exclude the marriage to the first birth spell because many are imputed.

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place,
sterilization of the woman or her husband,
or too few births are observed for the method to work.%
\footnote{
Most sterilizations take place right after a woman gives birth, and,
therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on the sex composition of prior children
with lower probabilities, the fewer boys. 
The effect is that the differences in parity progression probabilities are biased 
downwards.
}
For all spells, censoring is set at 96 months (eight years) after a woman can first
give birth in the spell.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.

@@

\subsection{Explanatory Variables}

@@

The models are estimated separately for each of the four education group to reduce any
potential problem from including other variables as proportional since
women with different education levels have different hazard profiles
\citep{Whitworth2002,Bhalotra2008,Kim2010}.
I capture sex composition with dummy variables for the
possible combinations, ignoring the ordering of births.
Area of residence is a dummy variable for living in an urban area.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell/education/period subsample and 
then predict average birth spacing, sex ratio, and the probability of having a birth 
in that spell.

@@

To find the expected average duration, I first calculate, for each woman, the probability 
of giving birth in each period $t$, which I use as weights to calculated her expected 
duration. 
I then average the individual expected durations across women using the parity progression
probabilities as weights.

The predicted parity progression probability is the likelihood of giving birth by 96 
months after the beginning of the spell.

The predicted sex ratio captures the percent of births that are boys for women in 
the sample when childbearing for that spell is over.
A woman's predicted sex ratio is the weighted average of the predicted percentage boys 
over each $t$ in the spell, calculated using equation (\ref{eq:probability_son}), using
the probability of giving birth at time $t$ as weights.%
\footnote{
With T=2, if 54 and 66 percent of the births are boys and the likelihoods of 
giving birth 20 and 40 percent, then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent of the
births are boys.
}
The predicted sex ratio shown is the weighted average of the individual predicted sex 
ratios, using the parity progression probabilities as weights.
For comparison, the graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

@@

\begin{sidewaysfigure}
\captionsetup[subfigure]{captionskip=-1pt,farskip=-0.5pt}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_urban_all}
    \end{minipage}
}
\caption{Estimated average spell length, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\ContinuedFloat
\rotatebox[origin=c]{90}{\small{Rural}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_rural_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_rural_all} 
    \end{minipage}
} 
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_rural_all} 
    \end{minipage}
} 
\caption{(Continued) Estimated average spell length, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}

@@

% \begin{figure}[htpb]
% \centering
% \rotatebox[origin=c]{90}{\small{Urban}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_med_urban_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_med_urban_all} 
%     \end{minipage}
% }
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_med_urban_all}
%     \end{minipage}
% }
% \\
% \rotatebox[origin=c]{90}{\small{Rural}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_med_rural_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_med_rural_all} 
%     \end{minipage}
% } 
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_med_rural_all} 
%     \end{minipage}
% } 
% \caption{Estimated average spell length, sex ratio, and probability of 
% a next birth for women with one to seven years of education}
% \label{fig:bootstrap_med}
% \end{figure}

\begin{sidewaysfigure}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_med_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_med_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_med_urban_all}
    \end{minipage}
}
\caption{Estimated average spell length, sex ratio, and probability of 
a next birth for women with one to seven years of education}
\label{fig:bootstrap_med}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\ContinuedFloat
\rotatebox[origin=c]{90}{\small{Rural}}
\subfloat[Second Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_med_rural_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_med_rural_all} 
    \end{minipage}
} 
\subfloat[Fourth Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_med_rural_all} 
    \end{minipage}
} 
\caption{(Continued) Estimated average spell length, sex ratio, and probability of 
a next birth for women with one to seven years of education}
\label{fig:bootstrap_med}
\end{sidewaysfigure}

@@

% \begin{figure}[htpb]
% \centering
% \rotatebox[origin=c]{90}{\small{Urban}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_high_urban_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_high_urban_all} 
%     \end{minipage}
% }
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_high_urban_all}
%     \end{minipage}
% }
% \\
% \rotatebox[origin=c]{90}{\small{Rural}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_high_rural_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_high_rural_all} 
%     \end{minipage}
% } 
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_high_rural_all} 
%     \end{minipage}
% } 
% \caption{Estimated average spell length, sex ratio, and probability of 
% a next birth for women with eight to eleven years of education}
% \label{fig:bootstrap_high}
% \end{figure}

\begin{sidewaysfigure}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_high_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_high_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_high_urban_all}
    \end{minipage}
}
\caption{Estimated average spell length, sex ratio, and probability of 
a next birth for women with eight to eleven years of education}
\label{fig:bootstrap_high}
\end{sidewaysfigure}

>>>>

First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up that captures both the non-randomness of
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, even without sex selection it is unlikely that characteristics, such as the sex
composition of previous births, have the same effects throughout the entire birth interval.
That is, however, the assumption behind the proportional hazard model, and if 
proportionality does not hold, the results are biased.
The proportionality assumption is especially problematic for higher-order birth 
intervals where there are substantial differences across groups both in the likelihood of 
progressing to the next birth and how soon couples want their next child if they are going 
to have one \citep{Whitworth2002,Bhalotra2008,Kim2010}.

The introduction of prenatal sex determination exacerbates any bias from
the proportionality assumption for two reasons. 
First, different groups have different levels of sex-selective abortion use and, thereby, 
birth spacing. 
Second, within a birth interval, a household's use of sex selection may vary and that
means that the effects of covariates vary as well.

I, therefore, use a non-proportional hazard specification that allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

The model is a discrete-time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell---the period from nine months after one birth to the 
following birth or censoring.
The spells begin nine months after the previous birth since this is the earliest we should 
expect to observe a new birth.
Censoring can happen for three reasons:
the survey takes place,
sterilization of the woman or her husband,
or there are too few births for the method to work.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ indicates which event took place.
The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise constant baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2).
\end{equation}
$D_j(t)$ is the piece-wise constant baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, and $Z_2$ captures the area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.%
\footnote{
With sex selection, the composition of prior children is, in principle, endogenous.
It is beyond the scope of this paper to develop a method for dealing with this issue.
}

% To further minimize any potential bias from assuming proportionality, estimations 
% are done separately for different levels of mothers' education and different periods.
% I discuss the choice of which variables to use for non-proportionality below.

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is a time period---here equal
to three months---I can estimate the model using a standard multinomial logit model.
%
% \footnote{
% The multinomial model assumes that alternative exit states are stochastically independent
% (Independence of Irrelevant Alternatives or IIA), which rules out any individual-specific 
% unmeasured or unobservable factors that affect both the hazard of having a girl and the 
% hazard of having a boy.
% I, therefore, include a proxy for fecundity discussed in Section \ref{sec:data}.
% Also, the multivariate probit model can be used as an alternative because the IIA is not 
% imposed \citep{han90}.
% The results are substantially identical between these two models and
% available upon request.
% }
In the reorganized data the outcome variable is 0 if the woman does not have a child in a 
given interval (the base outcome), 1 if she gives birth to a son in that interval, and 2 
if she gives birth to a daughter in that interval.

@@

I use the model to predict birth interval measures, the probability of having a birth, 
and the sex ratio.%
\footnote{
The interpretation of competing risk model coefficients is challenging \citep{thomas96}.
}
The sex ratio is based on the estimated percentage of children born at each $t$ that are 
boys, $\hat{Y}$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100,
\label{eq:probability_son}
\end{equation}
where $b$ and $g$ are the predicted probabilities of having a boy and a girl, respectively,
conditional on the explanatory variables and not having had a child before $t$.
Combining the percentage of boys and the likelihood of exiting the spell with a birth
across all $t$ gives the predicted percent of births that are boys over the entire spell.%
\footnote{
With T=2, if 54 and 66 percent of the births are boys and the likelihoods of 
giving birth 20 and 40 percent, then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent of the
births are boys.
}
The predicted sex ratio shown is the weighted average of the individual predicted sex 
ratios, using the parity progression probabilities as weights.

@@

The data come from the four rounds of the NFHS
collected in 1992--1993, 1998--1999, 2005--2006, and 2015--2016.
The surveys are large: 89,777, 90,303, 124,385, and 699,686 women, respectively. 
NFHS-1 and NFHS-2 surveyed only ever-married women, while the two later surveys included 
never-married.

I focus on the three birth intervals starting from the first birth and ending with the 
fourth birth.
I exclude the interval from marriage to the first birth because many are imputed and,
with falling fertility, few women had five or more births, especially among the 
better-educated.

I restrict the sample to Hindus, who constitute about 80 percent of India’s population.
If the use of sex selection differs across groups, assuming that the
baseline hazard is the same leads to bias. 
Furthermore, the other groups are each too small to estimate different
baseline hazards for each and so different that combining them into one
group would not make sense.

Finally, I exclude visitors, as well as women in any of the following 
categories: never married; no gauna yet; married more than once; divorced; 
not living with husband; inconsistent age at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported giving birth 
before age 12, had a birth before marriage, or an interval between births of less than 
nine months.

In addition to a large number of women surveyed and the long period covered, a
benefit of the NFHS is that enumerators pay careful attention to the spacing 
between births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, remains a potential problem.
Recall error is heavily dependent on how long ago a woman was married, and I, therefore, 
drop women married 22 years or more based on the discussion in the Appendix.

To ensure that there are enough births for the method to work, I censor each spell at 96 
months (eight years) after a woman can first give birth in the spell, which is equivalent 
to 105 months after the birth of the prior child.
Less than one percent of observed births occur after the cut-off.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

@@

@@

I estimate a separate model for each birth interval, education group, and period 
combination to minimize any potential bias from including proportional variables.
I capture sex composition with dummy variables for the
possible combinations, ignoring the ordering of births.
Area of residence is a dummy variable for living in an urban area.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell, education, and period 
subsample and then predict average birth spacing, sex ratio, and the probability of having 
a birth in that spell.

@@

To find the expected average birth interval, I first calculate, for each woman, the 
probability of giving birth in each period $t$, which I use as weights to calculated her 
expected spell lengths. 
I then average the individual expected spell lengths across women using the parity 
progression probabilities as weights.
Finally, I add nine months to account for spell start to arrive at birth interval.

The predicted parity progression probability is the likelihood of giving birth by the
end of the spell.%
\footnote{
Most sterilizations take place right after a woman gives birth, and,
therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on the sex composition of prior children
with lower probabilities, the fewer boys. 
The effect is to bias downward the differences in parity progression probabilities.
}
The predicted sex ratio captures the percent of births that are boys for women in 
the sample when childbearing for that spell is over, calculated as shown above.
For comparison, the graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

@@

\begin{sidewaysfigure}
\captionsetup[subfigure]{captionskip=-1pt,farskip=-0.5pt}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_urban_all}
    \end{minipage}
}
\caption{Estimated average birth interval, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\ContinuedFloat
\rotatebox[origin=c]{90}{\small{Rural}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_rural_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_rural_all} 
    \end{minipage}
} 
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_rural_all} 
    \end{minipage}
} 
\caption{(Continued) Estimated average birth interval, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}

@@

% \begin{figure}[htpb]
% \centering
% \rotatebox[origin=c]{90}{\small{Urban}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_med_urban_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_med_urban_all} 
%     \end{minipage}
% }
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_med_urban_all}
%     \end{minipage}
% }
% \\
% \rotatebox[origin=c]{90}{\small{Rural}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_med_rural_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_med_rural_all} 
%     \end{minipage}
% } 
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_med_rural_all} 
%     \end{minipage}
% } 
% \caption{Estimated average birth interval, sex ratio, and probability of 
% a next birth for women with one to seven years of education}
% \label{fig:bootstrap_med}
% \end{figure}

\begin{sidewaysfigure}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_med_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_med_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_med_urban_all}
    \end{minipage}
}
\caption{Estimated average birth interval, sex ratio, and probability of 
a next birth for women with one to seven years of education}
\label{fig:bootstrap_med}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\centering
\ContinuedFloat
\rotatebox[origin=c]{90}{\small{Rural}}
\subfloat[Second Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_med_rural_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_med_rural_all} 
    \end{minipage}
} 
\subfloat[Fourth Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_med_rural_all} 
    \end{minipage}
} 
\caption{(Continued) Estimated average birth interval, sex ratio, and probability of 
a next birth for women with one to seven years of education}
\label{fig:bootstrap_med}
\end{sidewaysfigure}

@@

% \begin{figure}[htpb]
% \centering
% \rotatebox[origin=c]{90}{\small{Urban}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_high_urban_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_high_urban_all} 
%     \end{minipage}
% }
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_high_urban_all}
%     \end{minipage}
% }
% \\
% \rotatebox[origin=c]{90}{\small{Rural}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell2_high_rural_all}
%     \end{minipage}
% } 
% \subfloat[Third Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell3_high_rural_all} 
%     \end{minipage}
% } 
% \subfloat[Fourth Spell]{
%     \begin{minipage}{0.29\textwidth}
%         \includegraphics[width=\textwidth]{bs_spell4_high_rural_all} 
%     \end{minipage}
% } 
% \caption{Estimated average birth interval, sex ratio, and probability of 
% a next birth for women with eight to eleven years of education}
% \label{fig:bootstrap_high}
% \end{figure}

\begin{sidewaysfigure}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_high_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_high_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.29\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_high_urban_all}
    \end{minipage}
}
\caption{Estimated average birth interval, sex ratio, and probability of 
a next birth for women with eight to eleven years of education}
\label{fig:bootstrap_high}
\end{sidewaysfigure}
